---
title: "Getting Started with mvher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mvher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

**mvher** provides principled missing-value handling for entity resolution (ER)
and record linkage pipelines.  The core problem: standard string-similarity
metrics (Jaro-Winkler, Levenshtein) treat two empty strings as *perfectly
matching* (`sim = 1`), which sends a spurious "definitely a match" signal to
the classifier for pairs where both records lack a field.

`mvher` offers:

- **Eight field-comparison strategies** controlling how empty/NA fields are
  scored during pairwise comparison.
- **Seven feature-matrix imputation strategies** for filling residual NAs
  after feature extraction.
- **Four diagnostic functions** for measuring and classifying missingness.
- A **strategy recommendation engine** that suggests the best approach
  given your data profile.

```{r setup}
library(mvher)
```

---

## 1. Diagnosing missingness

Build a synthetic six-record dataset with structured missingness:

```{r make-data}
records <- data.frame(
  id        = 1:6,
  entity_id = c(1L, 1L, 2L, 2L, 3L, 3L),
  name      = c("Alice Smith", "Alice Smyth", NA,    "Bob Jones", "",      "Carol White"),
  city      = c("London",      "Londen",      "Paris", NA,         "Rome",  "Roma"),
  stringsAsFactors = FALSE
)
```

### Field-level summary

```{r field-summary}
mv_field_summary(records, c("name", "city"))
```

### Pair-level summary

```{r pair-summary}
pairs <- data.frame(id1 = c(1, 2, 3, 5), id2 = c(2, 4, 4, 6))
mv_pair_summary(records, pairs, c("name", "city"))
```

### Missingness mechanism

```{r mech}
mv_detect_mechanism(records, c("name", "city"))
```

### Entity-level ICC

High ICC indicates that whole entities tend to be missing (MNAR-like pattern),
which creates covariate shift under entity-disjoint train/test splits:

```{r icc}
mv_entity_correlation(records, c("name", "city"))
```

---

## 2. Choosing a strategy

### Strategy overview table

```{r strat-table}
tbl <- mv_strategy_table()
tbl[tbl$type == "field-comparison", c("strategy", "handles_mnar", "adds_features", "description")]
```

### Automatic recommendation

```{r recommend}
# Moderate missingness with high entity-level correlation under entity-disjoint split
rec <- mv_recommend(field_miss_rate = 0.25, entity_corr = 0.6, split_type = "entity")
print(rec)
```

---

## 3. Field-comparison strategies

All eight strategies are accessed through `mv_compare_fields()` (vector-level)
or `mv_pair_features()` (pair-level feature matrix).

```{r compare-demo}
xa <- c("Alice Smith", "Bob Jones", "",     "")
xb <- c("Alice Smyth", "Robert J.", "Carol", "")

# Default (semantically wrong: both-empty → sim=1)
mv_compare_fields(xa, xb, method = "current")$jw

# Neutral: both-empty → 0.5
mv_compare_fields(xa, xb, method = "neutral")$jw

# Zero: any-empty → 0
mv_compare_fields(xa, xb, method = "zero")$jw

# Asymmetric: both-empty=0, one-empty=0.5
mv_compare_fields(xa, xb, method = "asymmetric")$jw

# Fellegi-Sunter integer gamma levels
mv_compare_fields(xa, xb, method = "fellegi_sunter")$gamma
```

### Building a feature matrix

```{r pair-features}
feats_neutral <- mv_pair_features(records, pairs,
                                  text_cols = c("name", "city"),
                                  method    = "neutral")
feats_neutral$X  # JW + LV for each column, 4 features total
feats_neutral$y  # match labels

feats_full <- mv_pair_features(records, pairs,
                               text_cols = c("name", "city"),
                               method    = "full")
names(feats_full$X)  # includes _any_miss and _both_miss indicator columns
```

---

## 4. Feature-matrix imputation

After extracting features (especially with `"complete_cases"` or `"indicator"`
methods that can leave NAs), use the imputer to fill residual missing values.

The two-step API (fit on training, apply to both) prevents data leakage:

```{r impute-two-step}
# Simulate a feature matrix with some residual NAs
Xtr <- data.frame(jw = c(0.9, NA, 0.3, 0.8), lv = c(NA, 0.7, 0.5, NA))
Xte <- data.frame(jw = c(NA, 0.6),            lv = c(0.4, NA))

# Fit on training data only
imp <- mv_fit_imputer(Xtr, strategy = "mean")
print(imp)

# Apply to both
Xtr_clean <- mv_apply_imputer(Xtr, imp)
Xte_clean <- mv_apply_imputer(Xte, imp)  # uses training means, not test means
Xtr_clean
Xte_clean
```

Or use the one-step convenience wrapper:

```{r impute-one-step}
res <- mv_impute(Xtr, Xte, strategy = "median")
res$X_train
res$X_test
```

### Available strategies

```{r strat-impute}
mv_strategy_table()[mv_strategy_table()$type == "imputation",
                    c("strategy", "handles_mar", "computation_cost", "description")]
```

---

## 5. End-to-end mini-example

```{r e2e}
# Records: 8 entities x 3 records each, some missing names
set.seed(42)
n <- 24
d <- data.frame(
  id        = seq_len(n),
  entity_id = rep(1:8, each = 3L),
  name      = ifelse(runif(n) < 0.2, NA_character_,
                     paste0(sample(c("Smith","Jones","Brown","White"), n, TRUE),
                            " ",
                            sample(c("Alice","Bob","Carol","Dan"), n, TRUE))),
  city      = ifelse(runif(n) < 0.15, "", sample(c("London","Paris","Rome"), n, TRUE)),
  stringsAsFactors = FALSE
)

# Step 1: Diagnose
cat("Field missingness:\n")
print(mv_field_summary(d, c("name", "city")))

# Step 2: Recommend strategy
rec <- mv_recommend(
  field_miss_rate = c(0.20, 0.15),
  entity_corr     = NULL,
  split_type      = "entity"
)
cat("\nRecommendation:", rec$field_method, "/", rec$imputer_strategy, "\n")

# Step 3: Train/test entity split (70/30)
set.seed(1)
ents     <- unique(d$entity_id)
test_ent <- sample(ents, floor(0.3 * length(ents)))
d_tr     <- d[!d$entity_id %in% test_ent, ]
d_te     <- d[ d$entity_id %in% test_ent, ]

# Step 4: Make all candidate pairs within each split
make_all_pairs <- function(ids) {
  ids <- sort(unique(ids)); n <- length(ids)
  if (n < 2) return(data.frame(id1 = integer(), id2 = integer()))
  data.frame(id1 = rep(ids[-n], times = rev(seq_len(n - 1))),
             id2 = unlist(lapply(2:n, function(i) ids[i:n])))
}
pairs_tr <- make_all_pairs(d_tr$id)
pairs_te <- make_all_pairs(d_te$id)

# Step 5: Extract features with recommended method
ftr <- mv_pair_features(d_tr, pairs_tr, c("name", "city"), method = rec$field_method)
fte <- mv_pair_features(d_te, pairs_te, c("name", "city"), method = rec$field_method)

# Step 6: Impute feature matrices
res <- mv_impute(ftr$X, fte$X, strategy = rec$imputer_strategy)

cat("\nTrain feature matrix (first 3 rows):\n")
print(head(res$X_train, 3))
cat("\nTrain label distribution:", table(ftr$y), "\n")
```

---

## Summary of strategy selection rules

| Condition | Recommended `method` | Recommended `strategy` |
|---|---|---|
| < 2% missingness | `"current"` | `"mean"` |
| Entity-disjoint split + ICC > 0.3 | `"full"` | `"mean"` |
| ≥ 15% missingness | `"neutral"` | `"mean"` |
| Unknown / low missingness | `"neutral"` | `"mean"` |
| Tree-based model | `"indicator"` | `"none"` |
| Strict Fellegi-Sunter model | `"fellegi_sunter"` | `"none"` |

See `mv_strategy_table()` for a complete comparison of all 15 strategies.
